---
title: "Virginia Invasive Plant Species Analysis"
author: "Ray Santiago and Seun Oladipo"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: 3
    df_print: paged
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

**What has been done**

1. Packages
Loaded tidyverse, readxl, scales, patchwork, sf, terra, tmap, geodata, dismo, randomForest, and lubridate. Two critical environment flags set at startup — sf_use_s2(FALSE) and OGR_ORGANIZE_POLYGONS = "SKIP" — to prevent polygon geometry slowdowns. Namespace conflicts manually resolved by assigning dplyr::select, dplyr::filter, dplyr::rename, and terra::extract explicitly, since raster/dismo packages mask them.

2. VIPS Spreadsheet Cleaning
Read vips-list-oct2024.xlsx (skip=1). Binary-encoded nine habitat/ecology columns (Mountain, Piedmont, Coastal, etc.) as integer flags. Trimmed whitespace from the VA Invasiveness Rank field. Dropped rows missing Common Name or Scientific Name. Output: 90 species retained in vips_clean.

3. Load Geodatabase
Built a named lifeform lookup vector for 18 species. Constructed a Virginia bounding box (xmin=-83.7, ymin=36.5, xmax=-75.2, ymax=39.5) with a 0.2° buffer, converted to WKT for wkt_filter. Read Bio_InvasivePlantCurrent.gdb/ with that filter — this was the critical fix that brought the raw load from 945,042 records (full national dataset) down to 44,682 records in the Virginia region.

4. Invasive Clean
Converted all polygons to centroids immediately at load — this was the core performance fix that solved all the hanging problems. Dropped sf geometry entirely (st_drop_geometry) so everything downstream runs as a plain dataframe. Extracted lon/lat as numeric columns, applied the hard bbox filter, selected the 8 needed columns, standardized two messy species name variants (Elaeagnus umbellata, Polygonum cespitosum) with str_detect, deduplicated on species + coordinate (fast, no geometry hashing), imputed missing lifeforms from the lookup vector, and median-imputed GIS_ACRES and DISTANCE_TO_WATER_HORIZONTAL. Output: 35,375 records after cleaning.

5. Clip to Virginia
Downloaded Virginia boundary via geodata::gadm("USA", level=1), filtered to NAME_1 == "Virginia", transformed to EPSG:4326. Rebuilt an sf object from the lon/lat columns, clipped with st_filter (not st_intersection, per the handoff), then immediately dropped geometry again. Retained invasive_va_sf as a named object for the county spatial join later. Output: 818 records within Virginia.

6. Focal Species
Filtered to the three focal species representing distinct lifeform classes: Ailanthus altissima (Tree, n=382), Microstegium vimineum (Graminoid, n=12), Rosa multiflora (Shrub, n=63). Microstegium flagged as data-limited at this stage — n=12 is insufficient for reliable SDM.

8. Full Analysis

8b. Background Points — Sampled 10,000 random background points from the Virginia climate raster using terra::spatSample with na.rm=TRUE so only land cells are sampled.
8c. Random Forest SDM — For each species: combined presences with background points (0/1), balanced class sampling (sampsize = c("0"=n, "1"=n)), trained Random Forest with 300 trees and importance=TRUE. Results: Ailanthus OOB accuracy 91.6%, Rosa 92.6%, Microstegium 88.7% (not interpretable, flagged).

8d. Variable Importance — Extracted Mean Decrease Accuracy via importance(type=1). For Ailanthus (the only reliable model): Annual Precip bio12=25.85, Temp Seasonality bio4=24.76, Mean Temp bio1=17.94, Precip Seasonality bio15=6.59. Rosa bio4 returned −15.98 (negative MDA = variable adding noise, likely due to low n=63).

8e. Statistical Tests — Kruskal-Wallis for temperature: χ²=16.623, df=2, p=0.0002. Kruskal-Wallis for precipitation: χ²=86.782, df=2, p<2.2×10⁻¹⁶. Pairwise Wilcoxon (Bonferroni): temperature — Ailanthus vs Microstegium p=0.007, Ailanthus vs Rosa p=0.013, Microstegium vs Rosa p=0.100 (n.s.). Precipitation — Ailanthus vs Rosa p<2×10⁻¹⁶, Microstegium vs Rosa p=0.032, Ailanthus vs Microstegium p=1.000 (n.s.).

8f. County Hotspots — Spatial join of invasive_va_sf against GADM Level 2 Virginia counties. Top 15 by record count, all in western Appalachian Virginia: Botetourt (68), Amherst (67), Rockbridge (62), Rockingham (61), Scott (61), Wise (60), Craig (59), Bath (52), Giles (52), Augusta (33), Shenandoah (32), Wythe (32), Grayson (30), Lee (30), Montgomery (26).







# 1. Packages

```{r packages}
library(tidyverse)
library(readxl)
library(scales)
library(patchwork)
library(sf)
library(terra)
library(tmap)
library(geodata)
library(dismo)
library(randomForest)
library(lubridate)

# Suppress geometry slowdown
sf_use_s2(FALSE)
Sys.setenv(OGR_ORGANIZE_POLYGONS = "SKIP")

# # Explicit conflict resolution — must come AFTER all library() calls
# conflictRules("dplyr", prefer = c("select", "filter", "rename"))
# options(conflicts.policy = list(warn = FALSE))

# Force-prefix the ones that matter
select <- dplyr::select
filter <- dplyr::filter
rename <- dplyr::rename
extract <- terra::extract
```

---

# 2. VIPS Spreadsheet Cleaning

```{r vips-clean}
vips_raw <- read_excel("vips-list-oct2024.xlsx", skip = 1)

vips_clean <- vips_raw %>%
  dplyr::mutate(across(
    c(Mountain, Piedmont, Coastal, Full, Partial, Shade, Hydric, Mesic, Xeric),
    ~ as.integer(!is.na(.)))) %>%
  dplyr::mutate(`VA Invasiveness Rank` = str_trim(`VA Invasiveness Rank`)) %>%
  drop_na(`Common Name`, `Scientific Name`)

cat("VIPS species:", nrow(vips_clean), "\n")
```

---

# 3. Load Geodatabase

```{r load-gdb}
# Lifeform lookup vector
lifeform_lookup <- c(
  "Ailanthus altissima"    = "Tree",       "Microstegium vimineum"  = "Graminoid",
  "Rosa multiflora"        = "Shrub",      "Elaeagnus umbellata"    = "Shrub",
  "Berberis thunbergii"    = "Shrub",      "Alliaria petiolata"     = "Forb/Herb",
  "Lonicera japonica"      = "Vine",       "Lonicera maackii"       = "Shrub",
  "Lonicera morrowii"      = "Shrub",      "Paulownia tomentosa"    = "Tree",
  "Celastrus orbiculatus"  = "Vine",       "Ligustrum vulgare"      = "Shrub",
  "Lespedeza cuneata"      = "Forb/Herb",  "Phalaris arundinacea"   = "Graminoid",
  "Tussilago farfara"      = "Forb/Herb",  "Leucanthemum vulgare"   = "Forb/Herb",
  "Securigera varia"       = "Forb/Herb",  "Polygonum cespitosum"   = "Forb/Herb"
)

# Virginia bounding box + small buffer for wkt_filter
va_bbox_buf <- st_bbox(
  c(xmin = -83.7, ymin = 36.5, xmax = -75.2, ymax = 39.5),
  crs = st_crs(4326)
) %>%
  st_as_sfc() %>%
  st_buffer(0.2) %>%
  st_as_text()

# Load with bbox filter — keeps ~40k instead of 945k
invasive_sf <- st_read("Bio_InvasivePlantCurrent.gdb/",
                        wkt_filter = va_bbox_buf,
                        quiet = TRUE)

cat("Raw records loaded:", nrow(invasive_sf), "\n")
```

---

# 4. Invasive Clean

```{r invasive-clean}
# KEY OPTIMIZATION: convert polygons → centroids immediately
# Eliminates all expensive polygon geometry ops downstream
invasive_pts <- invasive_sf %>%
  st_centroid() %>%          # polygons → single points; fast
  st_transform(crs = 4326)

rm(invasive_sf); gc()        # free ~500MB of polygon memory

# Extract lon/lat as plain columns
coords <- st_coordinates(invasive_pts)

invasive_va <- invasive_pts %>%
  st_drop_geometry() %>%     # now just a regular dataframe — no more sf overhead
  dplyr::mutate(
    lon = coords[, 1],
    lat = coords[, 2]
  ) %>%
  # Hard bbox filter
  dplyr::filter(
    !is.na(lon), !is.na(lat),
    between(lon, -83.7, -75.2),
    between(lat,  36.5,  39.5)
  ) %>%
  # Select columns
  dplyr::select(
    ACCEPTED_SCIENTIFIC_NAME, ACCEPTED_COMMON_NAME,
    LIFEFORM_DOMINANT, GIS_ACRES, DATE_COLLECTED,
    DISTANCE_TO_WATER_HORIZONTAL, lon, lat
  ) %>%
  # Standardise messy species names
  dplyr::mutate(ACCEPTED_SCIENTIFIC_NAME = case_when(
    str_detect(ACCEPTED_SCIENTIFIC_NAME, "Elaeagnus umbellata")  ~ "Elaeagnus umbellata",
    str_detect(ACCEPTED_SCIENTIFIC_NAME, "Polygonum cespitosum") ~ "Polygonum cespitosum",
    TRUE ~ ACCEPTED_SCIENTIFIC_NAME
  )) %>%
  # Fast dedup on name + coordinates (no geometry hashing needed)
  dplyr::distinct(ACCEPTED_SCIENTIFIC_NAME, lon, lat, .keep_all = TRUE) %>%
  # Fill missing lifeforms from lookup
  dplyr::mutate(LIFEFORM_DOMINANT = case_when(
    !is.na(LIFEFORM_DOMINANT)                             ~ LIFEFORM_DOMINANT,
    ACCEPTED_SCIENTIFIC_NAME %in% names(lifeform_lookup)  ~ lifeform_lookup[ACCEPTED_SCIENTIFIC_NAME],
    TRUE                                                  ~ "Unknown"
  )) %>%
  # Impute missing numeric fields with median
  dplyr::mutate(
    GIS_ACRES = ifelse(is.na(GIS_ACRES),
                       median(GIS_ACRES, na.rm = TRUE), GIS_ACRES),
    DISTANCE_TO_WATER_HORIZONTAL = ifelse(
      is.na(DISTANCE_TO_WATER_HORIZONTAL),
      median(DISTANCE_TO_WATER_HORIZONTAL, na.rm = TRUE),
      DISTANCE_TO_WATER_HORIZONTAL)
  )

rm(invasive_pts); gc()
cat("Records after cleaning:", nrow(invasive_va), "\n")
```

---

# 5. Clip to Virginia

```{r clip-to-virginia}
va_boundary <- geodata::gadm("USA", level = 1, path = "data/") %>%
  st_as_sf() %>%
  dplyr::filter(NAME_1 == "Virginia") %>%
  st_transform(crs = 4326)

# Rebuild sf from lon/lat, clip, then drop geometry again
invasive_va_sf <- invasive_va %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_filter(va_boundary)

# Keep as plain dataframe going forward (faster for all tabular ops)
invasive_va <- invasive_va_sf %>% st_drop_geometry()

cat("Records within Virginia:", nrow(invasive_va), "\n")
# Expect ~822
```

---

# 6. Focal Species

```{r focal-species}
focal_species <- c(
  "Ailanthus altissima",
  "Microstegium vimineum",
  "Rosa multiflora"
)

invasive_focal_df <- invasive_va %>%
  dplyr::filter(ACCEPTED_SCIENTIFIC_NAME %in% focal_species)

# Counts — Microstegium n=12 is expected and data-limited
invasive_focal_df %>%
  dplyr::group_by(ACCEPTED_SCIENTIFIC_NAME) %>%
  dplyr::summarise(
    n         = n(),
    lifeform  = dplyr::first(LIFEFORM_DOMINANT),
    mean_acres = round(mean(GIS_ACRES, na.rm = TRUE), 1)
  )
```

---

# 7. Climate Data

```{r climate-load}
va_clim_raw <- worldclim_tile(
  var  = "bio",
  lon  = -79,
  lat  = 38,
  res  = 0.5,
  path = "data/"
)

# Read tile directly and crop to Virginia extent
bio_path    <- sources(va_clim_raw[[1]])[1]
va_clim_crop <- rast(bio_path) %>%
  crop(ext(-83.7, -75.2, 36.5, 39.5))

# Keep only the 4 focal layers
va_clim_crop <- va_clim_crop[[c(1, 4, 12, 15)]]
names(va_clim_crop) <- c("bio1_mean_temp", "bio4_temp_seasonality",
                          "bio12_annual_precip", "bio15_precip_seasonality")

cat("Climate layers:", paste(names(va_clim_crop), collapse = ", "), "\n")
cat("Extent:", as.character(ext(va_clim_crop)), "\n")
```

---

# 8. Full Analysis

## 8a. Climate Extraction

```{r climate-extract}
pred_cols <- names(va_clim_crop)

focal_pts <- invasive_focal_df %>%
  dplyr::select(lon, lat) %>%
  vect(geom = c("lon", "lat"), crs = "EPSG:4326")

climate_vals <- terra::extract(va_clim_crop, focal_pts, ID = FALSE)

invasive_focal_df <- dplyr::bind_cols(invasive_focal_df, climate_vals) %>%
  dplyr::filter(complete.cases(dplyr::select(., all_of(pred_cols))))

cat("Focal records with climate data:", nrow(invasive_focal_df), "\n")

# Climate niche summary — values already in C/mm, NO /10
invasive_focal_df %>%
  dplyr::group_by(ACCEPTED_SCIENTIFIC_NAME) %>%
  dplyr::summarise(
    n              = n(),
    mean_temp_C    = round(mean(bio1_mean_temp,        na.rm = TRUE), 1),
    sd_temp        = round(sd(bio1_mean_temp,          na.rm = TRUE), 2),
    mean_precip_mm = round(mean(bio12_annual_precip,   na.rm = TRUE), 0),
    sd_precip      = round(sd(bio12_annual_precip,     na.rm = TRUE), 0)
  )
```

## 8b. Background Points

```{r background-pts}
set.seed(2024)
bg_pts <- terra::spatSample(
  va_clim_crop,
  size   = 10000,
  method = "random",
  xy     = TRUE,
  na.rm  = TRUE
) %>%
  as.data.frame() %>%
  dplyr::rename(lon = x, lat = y)

cat("Background points:", nrow(bg_pts), "\n")
```

## 8c. Random Forest SDM

```{r rf-sdm}
rf_models <- lapply(focal_species, function(sp) {
  pres <- invasive_focal_df %>%
    dplyr::filter(ACCEPTED_SCIENTIFIC_NAME == sp) %>%
    dplyr::select(all_of(pred_cols)) %>%
    dplyr::mutate(presence = 1L)

  bg <- bg_pts %>%
    dplyr::select(all_of(pred_cols)) %>%
    dplyr::mutate(presence = 0L)

  df <- dplyr::bind_rows(pres, bg) %>%
    dplyr::mutate(presence = as.factor(presence))

  n <- sum(df$presence == 1)

  randomForest(
    presence ~ .,
    data       = df,
    ntree      = 300,
    importance = TRUE,
    sampsize   = c("0" = n, "1" = n)
  )
})
names(rf_models) <- focal_species

# Accuracy
cat("\n=== OOB Accuracy ===\n")
for (sp in focal_species) {
  oob <- rf_models[[sp]]$err.rate[300, "OOB"]
  note <- if (sp == "Microstegium vimineum") " *** DATA-LIMITED n=12 ***" else ""
  cat(sp, "| Accuracy:", round(1 - oob, 3), note, "\n")
}
```

## 8d. Variable Importance

```{r variable-importance}
var_labels <- c(
  "bio1_mean_temp"           = "Mean Temp (bio1)",
  "bio4_temp_seasonality"    = "Temp Seasonality (bio4)",
  "bio12_annual_precip"      = "Annual Precip (bio12)",
  "bio15_precip_seasonality" = "Precip Seasonality (bio15)"
)

vi_df <- do.call(rbind, lapply(focal_species, function(sp) {
  imp <- importance(rf_models[[sp]], type = 1)
  data.frame(
    species    = sp,
    variable   = var_labels[rownames(imp)],
    importance = imp[, "MeanDecreaseAccuracy"]
  )
})) %>%
  dplyr::arrange(species, desc(importance))

cat("\n=== Variable Importance ===\n")
print(vi_df)
```

## 8e. Statistical Tests

```{r stats}
cat("\n=== Kruskal-Wallis: Temperature ===\n")
kruskal.test(bio1_mean_temp ~ ACCEPTED_SCIENTIFIC_NAME,
             data = invasive_focal_df) %>% print()

cat("\n=== Kruskal-Wallis: Precipitation ===\n")
kruskal.test(bio12_annual_precip ~ ACCEPTED_SCIENTIFIC_NAME,
             data = invasive_focal_df) %>% print()

cat("\n=== Pairwise Wilcoxon: Temperature (Bonferroni) ===\n")
pairwise.wilcox.test(
  invasive_focal_df$bio1_mean_temp,
  invasive_focal_df$ACCEPTED_SCIENTIFIC_NAME,
  p.adjust.method = "bonferroni"
) %>% print()

cat("\n=== Pairwise Wilcoxon: Precipitation (Bonferroni) ===\n")
pairwise.wilcox.test(
  invasive_focal_df$bio12_annual_precip,
  invasive_focal_df$ACCEPTED_SCIENTIFIC_NAME,
  p.adjust.method = "bonferroni"
) %>% print()
```

## 8f. County Hotspots

```{r county-hotspots}
va_counties <- geodata::gadm("USA", level = 2, path = "data/") %>%
  st_as_sf() %>%
  dplyr::filter(NAME_1 == "Virginia") %>%
  st_transform(crs = 4326)

# Rebuild sf from the full Virginia dataset for spatial join
invasive_va_sf %>%
  st_join(va_counties) %>%
  st_drop_geometry() %>%
  dplyr::count(NAME_2, sort = TRUE) %>%
  dplyr::rename(county = NAME_2, n_invasive = n) %>%
  head(15) %>%
  print()
```

---

# **Potential Figures**

## Figure 1: Time Series, same as before still needs to be adjusted.


```{r figures}
# Define these first — used across all panels
invasive_va_df <- invasive_va %>% sf::st_drop_geometry()
lifeform_colours <- c("Tree"="darkgreen","Shrub"="#e67e22",
                      "Graminoid"="#f1c40f","Vine"="#8e44ad",
                      "Forb/Herb"="#e74c3c","Unknown"="grey60")

# Panel 1: Detections over time
p1 <- invasive_va %>%
  sf::st_drop_geometry() %>%
  dplyr::mutate(year = lubridate::year(DATE_COLLECTED)) %>%
  dplyr::filter(!is.na(year), year >= 2015, year <= 2024) %>%
  dplyr::count(year, LIFEFORM_DOMINANT) %>%
  ggplot(aes(x = year, y = n, colour = LIFEFORM_DOMINANT)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = lifeform_colours, name = "Lifeform") +
  scale_x_continuous(breaks = 2015:2024) +
  labs(title = "Panel 1: Invasive Plant Detections Over Time — Virginia",
       x = NULL, y = "# Occurrence Records") +
  theme_minimal(base_size = 11)

# Panel 2: Top 10 species
p2 <- invasive_va_df %>%
  dplyr::count(ACCEPTED_SCIENTIFIC_NAME, sort = TRUE) %>%
  head(10) %>%
  ggplot(aes(x = n, y = reorder(ACCEPTED_SCIENTIFIC_NAME, n), fill = n)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = "#9ecfa8", high = "#1c3a24") +
  scale_x_continuous(expand = c(0,0)) +
  labs(title = "Panel 2: Top 10 Invasive Species in Virginia",
       x = "# Occurrences", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(face = "italic"))


p3 <- vips_clean %>%
  tidyr::pivot_longer(c(Mountain, Piedmont, Coastal),
                      names_to = "region", values_to = "present") %>%
  dplyr::filter(present == 1, !is.na(`VA Invasiveness Rank`)) %>%
  ggplot(aes(x = region, fill = `VA Invasiveness Rank`)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("High"="red3","Medium"="#c9742a","Low"="#3a7d44"),
                    name = "Invasiveness\nRank") +
  labs(title = "Panel 3: Species by Region & Invasiveness Rank",
       x = NULL, y = "# Species") +
  theme_minimal(base_size = 11)

p1 / p2 / p3
```

## Figure 2: Map
```{r Figure 1}

invasive_va_df <- invasive_va %>% st_drop_geometry()

lifeform_colours <- c("Tree"="darkgreen","Shrub"="#e67e22",
                       "Graminoid"="#f1c40f","Vine"="#8e44ad",
                       "Forb/Herb"="#e74c3c","Unknown"="grey60")

fig1 <- ggplot() +
  geom_sf(data = va_boundary, fill = "#f0ede6", colour = "grey50", linewidth = 0.4) +
  geom_point(data = invasive_va_df,
             aes(x = lon, y = lat, colour = LIFEFORM_DOMINANT),
             size = 0.8, alpha = 0.6) +
  scale_colour_manual(values = lifeform_colours, name = "Lifeform") +
  coord_sf(xlim = c(-83.7,-75.2), ylim = c(36.5,39.5), expand = FALSE) +
  labs(title = "Figure 1: Invasive Plant Species Occurrences in Virginia",
       caption = "Source: USFS Bio_InvasivePlantCurrent GDB") +
  theme_minimal(base_size = 12) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))
print(fig1)
```


Figure 3 - top 15 species bar chart

```{r figure2}

fig2 <- invasive_va_df %>%
  dplyr::count(ACCEPTED_SCIENTIFIC_NAME, LIFEFORM_DOMINANT, sort = TRUE) %>%
  head(15) %>%
  ggplot(aes(x = n, y = reorder(ACCEPTED_SCIENTIFIC_NAME, n), fill = LIFEFORM_DOMINANT)) +
  geom_col() +
  scale_fill_manual(values = lifeform_colours, name = "Lifeform") +
  scale_x_continuous(expand = c(0,0)) +
  labs(title = "Figure 2: Top 15 Invasive Species by Occurrence Count",
       x = "Number of Records", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(face = "italic"))
print(fig2)
```

# Potential Figures

## Fig 1 — Detection Time Series

```{r fig1-timeseries, fig.width=12, fig.height=8}
invasive_ts <- invasive_va %>%
  dplyr::mutate(year = year(as.Date(DATE_COLLECTED))) %>%
  dplyr::filter(!is.na(year), year >= 1990, year <= 2024)

p1 <- invasive_ts %>%
  dplyr::count(year) %>%
  ggplot(aes(x = year, y = n)) +
  geom_line(color = "#2d6a4f", linewidth = 0.9) +
  geom_point(color = "#2d6a4f", size = 2) +
  labs(title = "All Species: Annual Detections",
       x = "Year", y = "Records") +
  theme_minimal(base_size = 11)

p2 <- invasive_ts %>%
  dplyr::count(year, LIFEFORM_DOMINANT) %>%
  ggplot(aes(x = year, y = n, color = LIFEFORM_DOMINANT)) +
  geom_line(linewidth = 0.8) +
  labs(title = "Detections by Lifeform",
       x = "Year", y = "Records", color = "Lifeform") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p3 <- invasive_ts %>%
  dplyr::filter(ACCEPTED_SCIENTIFIC_NAME %in% focal_species) %>%
  dplyr::count(year, ACCEPTED_SCIENTIFIC_NAME) %>%
  ggplot(aes(x = year, y = n, color = ACCEPTED_SCIENTIFIC_NAME)) +
  geom_line(linewidth = 0.8) +
  labs(title = "Focal Species Detections",
       x = "Year", y = "Records", color = "Species") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        legend.text = element_text(face = "italic"))

p1 / p2 / p3
```

## Fig 2 — Occurrence Map

```{r fig2-map, fig.width=10, fig.height=7}
tmap_mode("plot")

tm_shape(va_boundary) +
  tm_borders(col = "grey40", lwd = 1.2) +
  tm_shape(invasive_va_sf) +
  tm_dots(
    col   = "LIFEFORM_DOMINANT",
    size  = 0.07,
    alpha = 0.7,
    title = "Lifeform",
    palette = "Set2"
  ) +
  tm_layout(
    main.title = "Invasive Plant Occurrences in Virginia",
    legend.outside = TRUE
  )
```

## Fig 3 — Top 15 Species

```{r fig3-top15, fig.width=10, fig.height=6}
invasive_va %>%
  dplyr::count(ACCEPTED_SCIENTIFIC_NAME, LIFEFORM_DOMINANT, sort = TRUE) %>%
  head(15) %>%
  dplyr::mutate(ACCEPTED_SCIENTIFIC_NAME = fct_reorder(ACCEPTED_SCIENTIFIC_NAME, n)) %>%
  ggplot(aes(x = n, y = ACCEPTED_SCIENTIFIC_NAME, fill = LIFEFORM_DOMINANT)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Top 15 Invasive Plant Species in Virginia",
    x = "Number of Records", y = NULL, fill = "Lifeform"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(face = "italic"))
```

## Fig 4 — Habitat Suitability Maps (Ailanthus + Rosa only)

```{r fig4-suitability, fig.width=12, fig.height=5}
# Predict over Virginia raster extent
plot_species <- c("Ailanthus altissima", "Rosa multiflora")

suit_maps <- lapply(plot_species, function(sp) {
  pred_df <- as.data.frame(va_clim_crop, xy = TRUE, na.rm = TRUE)
  pred_df$suitability <- predict(
    rf_models[[sp]],
    newdata = pred_df[, pred_cols],
    type    = "prob"
  )[, "1"]

  pred_rast <- rast(
    pred_df[, c("x", "y", "suitability")],
    type = "xyz",
    crs  = "EPSG:4326"
  )

  tm_shape(pred_rast) +
    tm_raster(
      col     = "suitability",
      palette = "YlOrRd",
      title   = "Suitability",
      style   = "cont",
      breaks  = seq(0, 1, 0.2)
    ) +
    tm_shape(va_boundary) +
    tm_borders(col = "grey30", lwd = 0.8) +
    tm_layout(
      main.title     = sp,
      main.title.fontface = "italic",
      legend.outside = TRUE
    )
})

tmap_arrange(suit_maps[[1]], suit_maps[[2]], ncol = 2)
```

## Fig 5 — Variable Importance (Ailanthus + Rosa only)

```{r fig5-varimp, fig.width=9, fig.height=5}
vi_df %>%
  dplyr::filter(species %in% plot_species) %>%
  dplyr::mutate(
    variable = factor(variable, levels = rev(unique(variable))),
    species  = factor(species, levels = plot_species)
  ) %>%
  ggplot(aes(x = importance, y = variable, fill = species)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c("Ailanthus altissima" = "#1b7837",
               "Rosa multiflora"     = "#c2185b"),
    labels = scales::label_wrap(20)
  ) +
  labs(
    title = "Variable Importance: Mean Decrease in Accuracy",
    x = "Mean Decrease Accuracy", y = NULL, fill = "Species"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.text = element_text(face = "italic"))
```
