---
title: "Virginia Invasive Plant Species Analysis"
author: "Ray Santiago and Seun Oladipo"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: 3
    df_print: paged
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE        # cache heavy chunks so re-knitting is fast
)
```

# 1. Libraries

```{r libraries}
library(tidyverse)
library(readxl)
library(scales)
library(patchwork)

# Spatial
library(sf)
library(terra)
library(tmap)

# Climate
library(geodata)

# SDM
library(dismo)        # Bioclim + background sampling
library(randomForest) # RF classifier for SDM
```

---

# 2. Data Loading & Cleaning

## 2a. VIPS spreadsheet

```{r vips}
vips_raw <- read_excel("vips-list-oct2024.xlsx", skip = 1)

vips_clean <- vips_raw |>
  mutate(
    # Presence flags: any non-NA = 1, NA = 0
    across(
      c(Mountain, Piedmont, Coastal, Full, Partial,
        Shade, Hydric, Mesic, Xeric),
      ~ as.integer(!is.na(.))
    ),
    # Strip trailing whitespace from rank (causes silent grouping errors)
    `VA Invasiveness Rank` = str_trim(`VA Invasiveness Rank`)
  ) |>
  drop_na(`Common Name`, `Scientific Name`)

glimpse(vips_clean)
```

## 2b. Invasive plant geodatabase

```{r invasive-load}
# Bounding box clipped to Virginia
va_bbox <- st_bbox(
  c(xmin = -83.7, ymin = 36.5, xmax = -75.2, ymax = 39.5),
  crs = st_crs(4326)
)

invasive_sf <- st_read(
  "Bio_InvasivePlantCurrent.gdb/",
  wkt_filter = st_as_text(st_as_sfc(va_bbox)),
  quiet = TRUE
)

cat("Raw records loaded:", nrow(invasive_sf), "\n")
```

```{r invasive-clean}
# Lookup table: fill missing lifeform from known species biology
lifeform_lookup <- c(
  "Ailanthus altissima"       = "Tree",
  "Microstegium vimineum"     = "Graminoid",
  "Rosa multiflora"           = "Shrub",
  "Elaeagnus umbellata"       = "Shrub",
  "Berberis thunbergii"       = "Shrub",
  "Alliaria petiolata"        = "Forb/Herb",
  "Lonicera japonica"         = "Vine",
  "Lonicera maackii"          = "Shrub",
  "Lonicera morrowii"         = "Shrub",
  "Paulownia tomentosa"       = "Tree",
  "Celastrus orbiculatus"     = "Vine",
  "Ligustrum vulgare"         = "Shrub",
  "Lespedeza cuneata"         = "Forb/Herb",
  "Phalaris arundinacea"      = "Graminoid",
  "Tussilago farfara"         = "Forb/Herb",
  "Leucanthemum vulgare"      = "Forb/Herb",
  "Securigera varia"          = "Forb/Herb",
  "Polygonum cespitosum"      = "Forb/Herb"
)

invasive_va <- invasive_sf |>
  # Step 1: Geometry validation
  filter(st_is_valid(SHAPE)) |>
  st_make_valid() |>

  # Step 2: Keep only needed columns
  select(
    ACCEPTED_SCIENTIFIC_NAME,
    ACCEPTED_COMMON_NAME,
    LIFEFORM_DOMINANT,
    INFESTED_AREA,
    GIS_ACRES,
    DATE_COLLECTED,
    OWNER_NAME,
    FS_UNIT_NAME,
    DISTANCE_TO_WATER_HORIZONTAL,
    SHAPE
  ) |>

  # Step 3: Collapse subspecies to species level
  mutate(
    ACCEPTED_SCIENTIFIC_NAME = case_when(
      str_detect(ACCEPTED_SCIENTIFIC_NAME, "Elaeagnus umbellata")  ~ "Elaeagnus umbellata",
      str_detect(ACCEPTED_SCIENTIFIC_NAME, "Polygonum cespitosum") ~ "Polygonum cespitosum",
      TRUE ~ ACCEPTED_SCIENTIFIC_NAME
    )
  ) |>

  # Step 4: Remove duplicates (same species x geometry = repeated survey)
  distinct(ACCEPTED_SCIENTIFIC_NAME, SHAPE, .keep_all = TRUE) |>

  # Step 5: Backfill lifeform and impute numeric NAs
  mutate(
    LIFEFORM_DOMINANT = case_when(
      !is.na(LIFEFORM_DOMINANT) ~ LIFEFORM_DOMINANT,
      ACCEPTED_SCIENTIFIC_NAME %in% names(lifeform_lookup) ~
        lifeform_lookup[ACCEPTED_SCIENTIFIC_NAME],
      TRUE ~ "Unknown"
    ),
    GIS_ACRES = replace_na(GIS_ACRES, median(GIS_ACRES, na.rm = TRUE)),
    DISTANCE_TO_WATER_HORIZONTAL = replace_na(
      DISTANCE_TO_WATER_HORIZONTAL,
      median(DISTANCE_TO_WATER_HORIZONTAL, na.rm = TRUE)
    )
  ) |>

  # Step 6: Extract polygon centroids for SDM point input
  mutate(
    lon = st_coordinates(st_centroid(SHAPE))[, 1],
    lat = st_coordinates(st_centroid(SHAPE))[, 2]
  ) |>
  filter(!is.na(lon), !is.na(lat)) |>

  # Step 7: Reproject to WGS84 (matches WorldClim)
  st_transform(crs = 4326)

invasive_df <- invasive_va |> st_drop_geometry()

cat("Records after cleaning:", nrow(invasive_df), "\n")
```

```{r cleaning-checks}
# Spot-check: coordinate bounds
invasive_df |>
  summarise(
    lon_min    = round(min(lon), 2),
    lon_max    = round(max(lon), 2),
    lat_min    = round(min(lat), 2),
    lat_max    = round(max(lat), 2),
    n_outside  = sum(lon < -83.7 | lon > -75.2 | lat < 36.5 | lat > 39.5)
  )

# Remaining NAs (OWNER_NAME / FS_UNIT_NAME are expected — not imputed)
invasive_df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "col", values_to = "na_count") |>
  filter(na_count > 0)
```

---

# 3. Climate Data

```{r climate}
va_clim <- worldclim_tile(var = "bio", lon = -79, lat = 38, res = 0.5, path = "data/")

# Crop to Virginia extent and select 4 predictors
va_clim_crop <- rast(sources(va_clim[[1]])[1]) |>
  crop(ext(-83.7, -75.2, 36.5, 39.5))

va_clim_crop <- va_clim_crop[[c(1, 4, 12, 15)]]
names(va_clim_crop) <- c("bio1_mean_temp", "bio4_temp_seasonality",
                          "bio12_annual_precip", "bio15_precip_seasonality")

cat("Climate layers:", paste(names(va_clim_crop), collapse = ", "), "\n")
```

---

# 4. Focal Species

```{r focal-species}
focal_species <- c(
  "Ailanthus altissima",   # Tree        — most records
  "Microstegium vimineum", # Graminoid   — shade-tolerant forest grass
  "Rosa multiflora"        # Shrub       — open-to-edge disturbed areas
)

# Filter and extract climate at centroid locations
focal_pts_vect <- invasive_df |>
  filter(ACCEPTED_SCIENTIFIC_NAME %in% focal_species) |>
  select(lon, lat) |>
  vect(geom = c("lon", "lat"), crs = "EPSG:4326")

climate_vals <- terra::extract(va_clim_crop, focal_pts_vect, ID = FALSE)

focal_df <- invasive_df |>
  filter(ACCEPTED_SCIENTIFIC_NAME %in% focal_species) |>
  bind_cols(climate_vals)

# Per-species summary
focal_df |>
  group_by(ACCEPTED_SCIENTIFIC_NAME) |>
  summarise(
    n                = n(),
    lifeform         = first(LIFEFORM_DOMINANT),
    mean_temp_C      = round(mean(bio1_mean_temp,        na.rm = TRUE) / 10, 1),
    annual_precip_mm = round(mean(bio12_annual_precip,   na.rm = TRUE), 0),
    temp_seasonality = round(mean(bio4_temp_seasonality, na.rm = TRUE), 0)
  )
```

---

# 5. Figures

## Figure 1: Top species occurrence counts + regional rank breakdown

```{r figure1, fig.height=8}
# ── Panel A: Top 10 species by record count ──────────────────────────────────
top10 <- invasive_df |>
  count(ACCEPTED_SCIENTIFIC_NAME, sort = TRUE) |>
  slice_head(n = 10) |>
  mutate(
    is_focal = ACCEPTED_SCIENTIFIC_NAME %in% focal_species,
    label    = str_replace(ACCEPTED_SCIENTIFIC_NAME, " ", "\n")
  )

p_counts <- ggplot(top10, aes(x = n, y = reorder(ACCEPTED_SCIENTIFIC_NAME, n),
                               fill = is_focal)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = comma(n)), hjust = -0.15, size = 3.2, color = "grey30") +
  scale_fill_manual(values = c("TRUE" = "#E57200", "FALSE" = "#4A6FA5")) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  scale_y_discrete(labels = function(x) str_replace(x, " ", "\n")) +
  labs(
    title    = "A  Top 10 invasive species by occurrence count",
    subtitle = "Orange = focal species selected for SDM",
    x        = "Occurrence records",
    y        = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold"),
    axis.text.y   = element_text(face = "italic", size = 9),
    panel.grid.major.y = element_blank()
  )

# ── Panel B: Species count by region and invasiveness rank ───────────────────
p_rank <- vips_clean |>
  pivot_longer(c(Mountain, Piedmont, Coastal),
               names_to = "Region", values_to = "Present") |>
  filter(Present == 1, !is.na(`VA Invasiveness Rank`)) |>
  count(Region, `VA Invasiveness Rank`) |>
  ggplot(aes(x = Region, y = n, fill = `VA Invasiveness Rank`)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = n), position = position_dodge(width = 0.9),
            vjust = -0.4, size = 3) +
  scale_fill_manual(
    values = c("High" = "#C0392B", "Medium" = "#E67E22", "Low" = "#27AE60"),
    name   = "Invasiveness\nRank"
  ) +
  labs(
    title = "B  Species count by physiographic region and severity rank",
    x     = NULL,
    y     = "Number of species"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

# ── Panel C: Lifeform composition ────────────────────────────────────────────
p_lifeform <- invasive_df |>
  filter(LIFEFORM_DOMINANT != "Unknown") |>
  count(LIFEFORM_DOMINANT, sort = TRUE) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = reorder(LIFEFORM_DOMINANT, n), y = n, fill = LIFEFORM_DOMINANT)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, accuracy = 1)), hjust = -0.15, size = 3.2) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  coord_flip() +
  labs(
    title = "C  Records by lifeform category",
    x     = NULL,
    y     = "Occurrence records"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

(p_counts | p_rank) / p_lifeform +
  plot_annotation(
    title   = "Figure 1. Virginia Invasive Plant Species — Dataset Overview",
    caption = "Source: USFS Bio_InvasivePlantCurrent.gdb; VIPS List Oct 2024",
    theme   = theme(plot.title = element_text(size = 13, face = "bold"))
  )
```

## Figure 2: Spatial distribution map

```{r figure2, fig.height=7}
# Virginia state boundary for context
va_state <- tigris::states(cb = TRUE, progress_bar = FALSE) |>
  filter(STUSPS == "VA") |>
  st_transform(4326)

# Hex-bin density map of all occurrence centroids
tmap_mode("plot")

tm_shape(va_state) +
  tm_borders(col = "grey40", lwd = 1.5) +
  tm_shape(invasive_va) +
  tm_dots(
    col       = "ACCEPTED_SCIENTIFIC_NAME",
    palette   = "Dark2",
    size      = 0.03,
    alpha     = 0.4,
    title     = "Species",
    legend.show = FALSE        # too many species; legend added as inset
  ) +
  tm_layout(
    main.title         = "Figure 2. Invasive Plant Distribution — Virginia National Forests (2019–2024)",
    main.title.size    = 0.95,
    main.title.fontface = "bold",
    frame              = FALSE,
    bg.color           = "white"
  ) +
  tm_compass(position = c("right", "top")) +
  tm_scale_bar(position = c("left", "bottom"))
```

## Figure 3: Climate niche comparison across focal species

```{r figure3, fig.height=7}
focal_long <- focal_df |>
  select(ACCEPTED_SCIENTIFIC_NAME,
         bio1_mean_temp, bio4_temp_seasonality,
         bio12_annual_precip, bio15_precip_seasonality) |>
  mutate(bio1_mean_temp = bio1_mean_temp / 10) |>   # convert to °C
  pivot_longer(
    -ACCEPTED_SCIENTIFIC_NAME,
    names_to  = "variable",
    values_to = "value"
  ) |>
  mutate(
    variable = recode(variable,
      bio1_mean_temp           = "BIO1: Mean Annual\nTemperature (°C)",
      bio4_temp_seasonality    = "BIO4: Temperature\nSeasonality",
      bio12_annual_precip      = "BIO12: Annual\nPrecipitation (mm)",
      bio15_precip_seasonality = "BIO15: Precipitation\nSeasonality"
    ),
    species_short = str_extract(ACCEPTED_SCIENTIFIC_NAME, "^\\w\\. \\w+|^\\w+") |>
      str_replace("Ailanthus", "A. altissima") |>
      str_replace("Microstegium", "M. vimineum") |>
      str_replace("Rosa", "R. multiflora"),
    species_short = case_when(
      str_detect(ACCEPTED_SCIENTIFIC_NAME, "Ailanthus")    ~ "A. altissima",
      str_detect(ACCEPTED_SCIENTIFIC_NAME, "Microstegium") ~ "M. vimineum",
      str_detect(ACCEPTED_SCIENTIFIC_NAME, "Rosa")         ~ "R. multiflora"
    )
  )

ggplot(focal_long, aes(x = species_short, y = value, fill = species_short)) +
  geom_violin(alpha = 0.6, trim = TRUE, color = NA) +
  geom_boxplot(width = 0.18, outlier.shape = NA, color = "grey20", fill = "white") +
  scale_fill_manual(values = c(
    "A. altissima"  = "#232D4B",
    "M. vimineum"   = "#E57200",
    "R. multiflora" = "#3a7d44"
  )) +
  facet_wrap(~ variable, scales = "free_y", nrow = 2) +
  labs(
    title    = "Figure 3. Bioclimatic Niche Profiles of Three Focal Invasive Species",
    subtitle = "Violin + boxplot of climate values at occurrence centroids",
    x        = NULL,
    y        = "Climate variable value",
    fill     = "Species",
    caption  = "Source: WorldClim 2.1 (0.5 arcmin); USFS occurrence centroids"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    strip.text       = element_text(face = "bold", size = 10),
    axis.text.x      = element_text(face = "italic", size = 9),
    legend.position  = "none"
  )
```

---

# 6. Species Distribution Model (SDM)

The SDM pipeline for each focal species:
1. **Occurrences** — centroid points, spatially thinned (1 pt per raster cell)
2. **Background** — 10,000 random points from within Virginia extent
3. **Predictors** — 4 WorldClim bioclimatic layers
4. **Model** — Random Forest classifier (presence vs. background)
5. **Output** — habitat suitability raster (0–1) for each species

```{r sdm-setup}
# Virginia raster mask (1 = inside VA, NA = outside)
va_vect  <- vect(va_state)
va_mask  <- terra::rasterize(va_vect, va_clim_crop[[1]], field = 1)

# Helper: thin occurrences to one point per raster cell
thin_to_cell <- function(pts_df, rast) {
  pts_vect   <- vect(pts_df, geom = c("lon", "lat"), crs = "EPSG:4326")
  cell_ids   <- cellFromXY(rast, crds(pts_vect))
  pts_df[!duplicated(cell_ids), ]
}

# Helper: generate background points within Virginia
make_background <- function(mask, n = 10000, seed = 42) {
  set.seed(seed)
  bg <- spatSample(mask, size = n, method = "random", na.rm = TRUE, xy = TRUE)
  names(bg)[1:2] <- c("lon", "lat")
  bg
}

# Extract predictors and label 1/0 for a single species
make_model_data <- function(species_name, focal_df, va_clim_crop, va_mask) {
  # Presence
  pres <- focal_df |>
    filter(ACCEPTED_SCIENTIFIC_NAME == species_name) |>
    thin_to_cell(va_clim_crop)

  pres_pts  <- vect(pres[, c("lon","lat")], geom = c("lon","lat"), crs="EPSG:4326")
  pres_vals <- terra::extract(va_clim_crop, pres_pts, ID = FALSE) |>
    mutate(presence = 1L)

  # Background
  bg        <- make_background(va_mask)
  bg_pts    <- vect(bg[, c("lon","lat")],  geom = c("lon","lat"), crs="EPSG:4326")
  bg_vals   <- terra::extract(va_clim_crop, bg_pts, ID = FALSE) |>
    mutate(presence = 0L)

  bind_rows(pres_vals, bg_vals) |>
    drop_na() |>
    mutate(presence = factor(presence, levels = c(0, 1)))
}
```

```{r sdm-run, cache=TRUE}
# Run RF SDM for each focal species
sdm_results <- map(focal_species, function(sp) {
  cat("Fitting SDM:", sp, "\n")

  model_data <- make_model_data(sp, focal_df, va_clim_crop, va_mask)

  set.seed(123)
  rf <- randomForest(
    presence ~ bio1_mean_temp + bio4_temp_seasonality +
               bio12_annual_precip + bio15_precip_seasonality,
    data       = model_data,
    ntree      = 500,
    importance = TRUE
  )

  # Predict suitability across Virginia
  suitability <- terra::predict(va_clim_crop, rf, type = "prob",
                                 index = 2, na.rm = TRUE) * va_mask

  list(species = sp, model = rf, suitability = suitability, data = model_data)
})

names(sdm_results) <- focal_species
```

## SDM diagnostics — variable importance

```{r sdm-importance, fig.height=5}
importance_df <- map_dfr(sdm_results, function(res) {
  imp <- importance(res$model, type = 1)   # mean decrease accuracy
  tibble(
    species  = res$species,
    variable = rownames(imp),
    MDA      = imp[, 1]
  )
}) |>
  mutate(
    variable = recode(variable,
      bio1_mean_temp           = "BIO1: Mean Temp",
      bio4_temp_seasonality    = "BIO4: Temp Seasonality",
      bio12_annual_precip      = "BIO12: Ann. Precip",
      bio15_precip_seasonality = "BIO15: Precip Seasonality"
    ),
    species_short = case_when(
      str_detect(species, "Ailanthus")    ~ "A. altissima",
      str_detect(species, "Microstegium") ~ "M. vimineum",
      str_detect(species, "Rosa")         ~ "R. multiflora"
    )
  )

ggplot(importance_df, aes(x = reorder(variable, MDA), y = MDA, fill = species_short)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ species_short, ncol = 3) +
  scale_fill_manual(values = c(
    "A. altissima"  = "#232D4B",
    "M. vimineum"   = "#E57200",
    "R. multiflora" = "#3a7d44"
  )) +
  labs(
    title   = "Variable Importance — Random Forest SDM",
    subtitle = "Mean Decrease in Accuracy (higher = more important predictor)",
    x = NULL, y = "Mean Decrease Accuracy"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title  = element_text(face = "bold"),
    strip.text  = element_text(face = "bold.italic")
  )
```

## Figure 4: Habitat Suitability Maps

```{r figure4, fig.height=8}
tmap_mode("plot")

# Build one panel per species
maps <- map(sdm_results, function(res) {
  sp_short <- case_when(
    str_detect(res$species, "Ailanthus")    ~ "Ailanthus altissima (Tree)",
    str_detect(res$species, "Microstegium") ~ "Microstegium vimineum (Graminoid)",
    str_detect(res$species, "Rosa")         ~ "Rosa multiflora (Shrub)"
  )

  tm_shape(res$suitability) +
    tm_raster(
      palette    = c("#f7f7f7", "#fee391", "#fd8d3c", "#bd0026"),
      n          = 9,
      style      = "cont",
      title      = "Suitability",
      breaks     = seq(0, 1, by = 0.2),
      colorNA    = NULL
    ) +
  tm_shape(va_state) +
    tm_borders(col = "grey30", lwd = 0.8) +
  tm_layout(
    main.title          = sp_short,
    main.title.size     = 0.85,
    main.title.fontface = "bold.italic",
    frame               = FALSE,
    legend.position     = c("right", "bottom"),
    legend.text.size    = 0.7
  )
})

tmap_arrange(maps[[1]], maps[[2]], maps[[3]], ncol = 1,
             outer.margins = 0.02) |>
  tmap_save_options() # comment out if running interactively

# Print with title
tmap_arrange(maps[[1]], maps[[2]], maps[[3]], ncol = 1)
```

---

# 7. Summary Table

```{r summary-table}
map_dfr(sdm_results, function(res) {
  # AUC proxy: OOB error from RF (1 - OOB is a rough accuracy)
  oob_acc  <- 1 - res$model$err.rate[nrow(res$model$err.rate), "OOB"]

  # Mean suitability across VA cells
  suit_vals <- values(res$suitability, na.rm = TRUE)

  tibble(
    Species            = res$species,
    N_occurrences      = sum(res$data$presence == 1),
    N_background       = sum(res$data$presence == 0),
    RF_OOB_accuracy    = round(oob_acc, 3),
    Mean_suitability   = round(mean(suit_vals), 3),
    High_suit_pct      = round(mean(suit_vals > 0.6) * 100, 1)
  )
}) |>
  rename(
    "Species"              = Species,
    "Occurrence N"         = N_occurrences,
    "Background N"         = N_background,
    "OOB Accuracy"         = RF_OOB_accuracy,
    "Mean Suitability"     = Mean_suitability,
    "% Highly Suitable (>0.6)" = High_suit_pct
  )
```
